---
title: "Understanding Causal Effects: Path Coefficients, Direct Effects, and Total Effects"
author: "Claude Sonnet 4.5 with Extended Thinking as Requested by Todd R. Johnson"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
---

## Introduction

This document demonstrates the key distinctions between:

- **Path coefficients** in linear structural equation models
- **Direct effects** (controlled and natural)
- **Total effects** and indirect effects
- How these concepts differ in linear vs. non-linear models
- The connection to causal DAGs

This version was produced by Claude Sonnet 4.5 with extended thinking. I have scanned the output, which appears to be correct; however, as a tutorial it is missing definitions and explanations. I will be embellishing it to provide a complete tutorial of these concepts.

Part of the reason for posting this is to demonstrate how to use LLMs to better understand a topic. To assess what Claude produced here and create a useful tutorial, I need to critically review the results and then fill in information that Claude left out. This requires active engagement on my part. Teaching and learning research shows that active engagement with material is one of the most important factors in learning. 

## Setup

```{r}
# Load required packages
library(tidyverse)
library(lavaan)
library(dagitty)
library(ggdag)
library(knitr)
library(kableExtra)

# Set seed for reproducibility
set.seed(123)
```

## Part 1: Linear Model (No Interaction)

### The Causal DAG

In this section, we'll work with a simple mediation model:
**X → M → Y** with a direct path **X → Y**

```{r}
# Create DAG
dag <- dagitty('dag {
  X -> M
  M -> Y
  X -> Y
}')

# Plot the DAG
ggdag(dag, layout = "circle") +
  theme_dag() +
  ggtitle("Simple Mediation DAG")
```

### Generate Data (Linear, No Interaction)

```{r}
n <- 5000

# Generate data according to structural equations
# M = 0.5*X + e_M
# Y = 0.3*X + 0.4*M + e_Y

data_linear <- tibble(
  X = rnorm(n, mean = 0, sd = 1),
  e_M = rnorm(n, mean = 0, sd = 1),
  e_Y = rnorm(n, mean = 0, sd = 1)
) %>%
  mutate(
    M = 0.5 * X + e_M,
    Y = 0.3 * X + 0.4 * M + e_Y
  )
```

### Path Analysis with lavaan

```{r}
# Specify the path model
model_linear <- '
  # Regressions
  M ~ a*X
  Y ~ b*M + c_prime*X
  
  # Define indirect and total effects
  indirect := a*b
  total := c_prime + a*b
  direct := c_prime
'

# Fit the model
fit_linear <- sem(model_linear, data = data_linear)

# Display results
summary(fit_linear, standardized = TRUE)
```

### Path Coefficients vs. Causal Effects

```{r}
# Extract parameter estimates
params_linear <- parameterEstimates(fit_linear) %>%
  filter(op %in% c("~", ":=")) %>%
  select(label, est, se, pvalue) %>%
  filter(!is.na(label))

kable(params_linear, 
      caption = "Path Coefficients and Effects in Linear Model",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key Point**: In the linear model **without interaction**:

- Path coefficient `c_prime` (X→Y) = Direct Effect = 0.3
- Path coefficient `a` (X→M) × path coefficient `b` (M→Y) = Indirect Effect = 0.5 × 0.4 = 0.2
- Total Effect = Direct + Indirect = 0.3 + 0.2 = 0.5

### Manual Calculation of Effects

```{r}
# Calculate effects manually using regression
m1 <- lm(M ~ X, data = data_linear)
m2 <- lm(Y ~ X + M, data = data_linear)
m3 <- lm(Y ~ X, data = data_linear)

results_linear <- tibble(
  Effect = c("a (X→M)", "b (M→Y|X)", "c' (Direct Effect)", 
             "a×b (Indirect Effect)", "c (Total Effect)"),
  Estimate = c(
    coef(m1)["X"],
    coef(m2)["M"],
    coef(m2)["X"],
    coef(m1)["X"] * coef(m2)["M"],
    coef(m3)["X"]
  )
)

kable(results_linear, 
      caption = "Manual Calculation of Effects",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Controlled Direct Effect (CDE)

In the linear model, the CDE is the same for all values of M:

```{r}
# The CDE(m) is the coefficient on X in the model Y ~ X + M
# It doesn't vary with m in a linear model

cde_values <- tibble(
  m = seq(-2, 2, by = 0.5),
  CDE = coef(m2)["X"]  # Same for all values of m
)

ggplot(cde_values, aes(x = m, y = CDE)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = coef(m2)["X"], linetype = "dashed", color = "red") +
  labs(title = "Controlled Direct Effect in Linear Model",
       subtitle = "CDE does not vary with m (no interaction)",
       x = "Value of Mediator (m)",
       y = "CDE(m)") +
  theme_minimal()
```

### Natural Direct and Indirect Effects

```{r}
# In linear models without interaction:
# NDE = CDE = path coefficient c'
# NIE = a × b

nde_linear <- coef(m2)["X"]
nie_linear <- coef(m1)["X"] * coef(m2)["M"]
total_linear <- nde_linear + nie_linear

cat("Natural Direct Effect (NDE):", round(nde_linear, 3), "\n")
cat("Natural Indirect Effect (NIE):", round(nie_linear, 3), "\n")
cat("Total Effect:", round(total_linear, 3), "\n")
cat("Sum of NDE + NIE:", round(nde_linear + nie_linear, 3), "\n")
```

**Important**: In linear models without interaction, CDE = NDE for all values of m.

---

## Part 2: Model with Interaction (Non-Linear Effects)

### Generate Data with Interaction

Now we add an X×M interaction term to create non-linearity:

```{r}
# Generate data with interaction
# M = 0.5*X + e_M
# Y = 0.3*X + 0.4*M + 0.2*X*M + e_Y

data_interaction <- tibble(
  X = rnorm(n, mean = 0, sd = 1),
  e_M = rnorm(n, mean = 0, sd = 1),
  e_Y = rnorm(n, mean = 0, sd = 1)
) %>%
  mutate(
    M = 0.5 * X + e_M,
    Y = 0.3 * X + 0.4 * M + 0.2 * X * M + e_Y
  )
```

### Fit Models with Interaction

```{r}
# Regression models
m1_int <- lm(M ~ X, data = data_interaction)
m2_int <- lm(Y ~ X * M, data = data_interaction)  # Includes interaction
m3_int <- lm(Y ~ X, data = data_interaction)

# Display regression with interaction
summary(m2_int)$coefficients %>%
  kable(caption = "Outcome Regression with Interaction",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Controlled Direct Effect (CDE) - Now Varies with m

With interaction, the CDE depends on the level of M:

```{r}
# Extract coefficients
theta0 <- coef(m2_int)["(Intercept)"]
theta1 <- coef(m2_int)["X"]
theta2 <- coef(m2_int)["M"]
theta3 <- coef(m2_int)["X:M"]

# CDE(m) = theta1 + theta3*m
# This is the effect of changing X by 1 unit when M is held at level m

m_values <- seq(-2, 2, by = 0.2)
cde_interaction <- tibble(
  m = m_values,
  CDE = theta1 + theta3 * m_values
)

ggplot(cde_interaction, aes(x = m, y = CDE)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = theta1, linetype = "dashed", color = "red",
             alpha = 0.5) +
  annotate("text", x = 0, y = theta1 + 0.05, 
           label = "Pure Direct Effect", color = "red") +
  labs(title = "Controlled Direct Effect with Interaction",
       subtitle = "CDE(m) = θ₁ + θ₃×m varies with the level of mediator",
       x = "Value of Mediator (m)",
       y = "CDE(m)") +
  theme_minimal()
```

**Key Insight**: With interaction, there is no single "direct effect" - it depends on where you fix the mediator!

### Natural Direct and Indirect Effects with Interaction

```{r}
# For binary X going from 0 to 1, with continuous M
# We'll use the formulas from VanderWeele (2015)

# Coefficients from regressions
beta0 <- coef(m1_int)["(Intercept)"]
beta1 <- coef(m1_int)["X"]  # a path

# Calculate E[M|X=0]
mean_M_X0 <- beta0

# Pure NDE (holds M at level when X=0)
# NDE_pure = (theta1 + theta3 * E[M|X=0])
nde_pure <- theta1 + theta3 * mean_M_X0

# Total NDE (holds M at level when X=1) 
mean_M_X1 <- beta0 + beta1
nde_total <- theta1 + theta3 * mean_M_X1

# Pure NIE (changes M from X=0 to X=1, keeps X at 0)
nie_pure <- (theta2 * beta1) + (theta3 * beta1 * mean_M_X0)

# Total NIE (changes M from X=0 to X=1, keeps X at 1)
nie_total <- (theta2 * beta1) + (theta3 * beta1 * mean_M_X1)

# Total effect
total_effect <- coef(m3_int)["X"]

effects_df <- tibble(
  Effect = c("Pure NDE", "Total NDE", "Pure NIE", "Total NIE", 
             "Pure NDE + Pure NIE", "Total NDE + Total NIE", "Total Effect"),
  Value = c(nde_pure, nde_total, nie_pure, nie_total,
            nde_pure + nie_pure, nde_total + nie_total, total_effect)
)

kable(effects_df, 
      caption = "Natural Direct and Indirect Effects with Interaction",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key Observations**:

1. Pure NDE + Pure NIE = Total Effect ✓
2. Total NDE + Total NIE = Total Effect ✓  
3. Pure NDE ≠ Total NDE (they differ in how they handle the X×M interaction)
4. Pure NIE ≠ Total NIE (they differ in how they handle the X×M interaction)
5. The "pure" effects attribute the interaction to the indirect effect
6. The "total" effects attribute the interaction to the direct effect

### Visualizing the Decomposition

```{r}
# Create a bar chart showing effect decomposition
decomp_data <- tibble(
  Type = rep(c("Pure Decomposition", "Total Decomposition"), each = 2),
  Component = rep(c("Direct Effect", "Indirect Effect"), 2),
  Value = c(nde_pure, nie_pure, nde_total, nie_total)
)

ggplot(decomp_data, aes(x = Type, y = Value, fill = Component)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_hline(yintercept = total_effect, linetype = "dashed", 
             color = "red", size = 1) +
  annotate("text", x = 1.5, y = total_effect + 0.03, 
           label = "Total Effect", color = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Natural Direct and Indirect Effects Decomposition",
       subtitle = "Both decompositions sum to the total effect",
       y = "Effect Size") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Part 3: Comparison Summary Table

```{r}
comparison <- tibble(
  Characteristic = c(
    "Path coefficients interpretable as causal effects?",
    "CDE varies with level of M?",
    "NDE = CDE?",
    "Single direct effect value?",
    "Indirect effect calculation",
    "Effect decomposition"
  ),
  `Linear Model (No Interaction)` = c(
    "Yes",
    "No - constant across all m",
    "Yes, they are equal",
    "Yes - one unique value",
    "a × b (product of path coefficients)",
    "TE = DE + IE (simple sum)"
  ),
  `Model with Interaction` = c(
    "No - effects are conditional",
    "Yes - CDE(m) = θ₁ + θ₃×m",
    "No - they differ",
    "No - depends on reference level",
    "Requires counterfactual formulas",
    "TE = Pure(NDE+NIE) = Total(NDE+NIE)"
  )
)

kable(comparison, 
      caption = "Key Differences: Linear vs. Non-Linear Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "15em") %>%
  column_spec(2:3, width = "15em")
```

---

## Part 4: Practical Example with Interpretation

Let's create a realistic example: effect of **education (X)** on **income (Y)** mediated by **job prestige (M)**.

```{r}
# Generate realistic data
set.seed(456)
n <- 2000

practical_data <- tibble(
  education = rnorm(n, mean = 14, sd = 2),  # Years of education
  e_prestige = rnorm(n, mean = 0, sd = 5),
  e_income = rnorm(n, mean = 0, sd = 8)
) %>%
  mutate(
    prestige = 3 * education + e_prestige,  # Job prestige score
    # Income depends on education, prestige, AND their interaction
    income = 2 * education + 1.5 * prestige + 0.3 * education * prestige + e_income
  )

# Visualize the data
ggplot(practical_data, aes(x = education, y = income, color = prestige)) +
  geom_point(alpha = 0.4) +
  scale_color_viridis_c() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Education, Job Prestige, and Income",
       x = "Years of Education",
       y = "Income (thousands)",
       color = "Job Prestige") +
  theme_minimal()
```

### Fit Models

```{r}
# Mediator model
med_model <- lm(prestige ~ education, data = practical_data)

# Outcome model with interaction
out_model <- lm(income ~ education * prestige, data = practical_data)

# Total effect model
tot_model <- lm(income ~ education, data = practical_data)
```

### Calculate and Interpret Effects

```{r}
# Extract coefficients
a_coef <- coef(med_model)["education"]
b0_coef <- coef(out_model)["(Intercept)"]
b1_coef <- coef(out_model)["education"]
b2_coef <- coef(out_model)["prestige"]
b3_coef <- coef(out_model)["education:prestige"]

# Mean prestige at different education levels
mean_prestige_low <- predict(med_model, 
                              newdata = data.frame(education = 12))
mean_prestige_high <- predict(med_model, 
                               newdata = data.frame(education = 16))

# CDE at different prestige levels
prestige_levels <- c(30, 40, 50, 60)
cde_values_practical <- sapply(prestige_levels, 
                               function(m) b1_coef + b3_coef * m)

cde_table <- tibble(
  `Prestige Level` = prestige_levels,
  `CDE (Effect per Year of Education)` = cde_values_practical,
  `Interpretation` = case_when(
    prestige_levels == 30 ~ "Low prestige job",
    prestige_levels == 40 ~ "Medium-low prestige",
    prestige_levels == 50 ~ "Medium-high prestige",
    prestige_levels == 60 ~ "High prestige job"
  )
)

kable(cde_table,
      caption = "Controlled Direct Effects at Different Prestige Levels",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation**: The direct effect of education on income INCREASES as we move to higher prestige jobs. This makes sense - education may be more valuable (in terms of income) in high-prestige positions.

### Natural Effects

```{r}
# Calculate natural effects for 1-year increase in education
mean_prestige_base <- predict(med_model, 
                               newdata = data.frame(education = 14))
mean_prestige_increased <- predict(med_model, 
                                    newdata = data.frame(education = 15))

# Pure NDE (fixes prestige at baseline education level)
pure_nde <- b1_coef + b3_coef * mean_prestige_base

# Pure NIE (changes prestige due to education, holds education at baseline)
pure_nie <- (b2_coef + b3_coef * 14) * a_coef

# Total effect
total_eff <- coef(tot_model)["education"]

interpretation_table <- tibble(
  Effect = c("Pure Natural Direct Effect",
             "Pure Natural Indirect Effect",
             "Total Effect",
             "Sum of Pure NDE + NIE"),
  Value = c(pure_nde, pure_nie, total_eff, pure_nde + pure_nie),
  `Interpretation` = c(
    "Income change from education, not through prestige",
    "Income change through increased prestige from education",
    "Total income change from 1 more year of education",
    "Sum of decomposed effects (should equal total)"
  )
)

kable(interpretation_table,
      caption = "Natural Effects Decomposition",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Proportion Mediated

```{r}
prop_mediated <- pure_nie / total_eff

cat("Proportion of total effect mediated by prestige:", 
    round(prop_mediated * 100, 1), "%\n")
```

---

## Part 5: Connection to DAGs and Assumptions

### Required Assumptions for Identification

```{r}
# Create a more complex DAG showing confounders
dag_confounded <- dagitty('dag {
  X -> M
  M -> Y
  X -> Y
  C -> X
  C -> Y
  C -> M
}')

tidy_dag <- tidy_dagitty(dag_confounded)

ggdag(tidy_dag) +
  theme_dag() +
  ggtitle("Mediation with Confounding") +
  geom_dag_text(color = "white") +
  geom_dag_edges() +
  geom_dag_node(aes(color = name == "C")) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "steelblue"),
                     guide = "none")
```

**Key Assumptions for Natural Direct/Indirect Effects**:

1. **No unmeasured X → Y confounding** (conditional on C)
2. **No unmeasured M → Y confounding** (conditional on C and X)
3. **No unmeasured X → M confounding** (conditional on C)
4. **No M → Y confounders affected by X** (this is the strong one!)

The fourth assumption is often violated in practice - this is why controlled direct effects are sometimes preferred (they only need assumptions 1-3).

---

## Summary and Key Takeaways

### 1. Linear Models (No Interaction)
- Path coefficients = Causal direct effects
- CDE = NDE = path coefficient on X→Y
- Simple decomposition: TE = DE + IE
- Effects are **transportable** across contexts

### 2. Models with Interaction
- Path coefficients ≠ Causal effects (effects are conditional)
- CDE varies with level of mediator M
- NDE requires counterfactual reasoning
- Two valid decompositions (pure vs. total)
- Effects are **context-dependent**

### 3. Practical Implications
- Always check for interactions before interpreting "direct effects"
- CDE is useful for policy questions: "What if we set M to m?"
- NDE is useful for mechanism questions: "What's the effect not through M?"
- Natural effects require stronger assumptions than controlled effects

### 4. Relationship to DAGs
- DAGs show causal structure (not necessarily linear relationships)
- Arrows represent direct causal effects
- d-separation helps identify confounding
- Same DAG can represent linear or non-linear models

## References

- Pearl, J. (2009). *Causality: Models, Reasoning, and Inference* (2nd ed.). Cambridge University Press.
- VanderWeele, T. J. (2015). *Explanation in Causal Inference: Methods for Mediation and Interaction*. Oxford University Press.
- Robins, J. M., & Greenland, S. (1992). Identifiability and exchangeability for direct and indirect effects. *Epidemiology*, 3, 143-155.
- VanderWeele, T. J., & Vansteelandt, S. (2009). Conceptual issues concerning mediation, interventions and composition. *Statistics and Its Interface*, 2, 457-468.

---

## Session Info

```{r}
sessionInfo()
```
