---
title: "Effect Modification vs. Interaction: A Simulation Study"
author: "Causal Inference Tutorial"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This document explores the distinction between **effect modification**
and **interaction** as defined by Hernán and Robins in their causal
inference framework. Through simulations and directed acyclic graphs
(DAGs), we'll demonstrate why these concepts, while related, represent
different causal questions.

# Key Definitions

## Effect Modification

**Effect modification** occurs when the effect of an intervention
(treatment) X on outcome Y varies across levels of another variable Z.

-   Z is called an **effect modifier**
-   Formally: The causal effect $E[Y^{x=1}] - E[Y^{x=0}]$ differs across
    strata of Z
-   Z is typically an *observed* (not manipulated) variable
-   Example: A drug's effect differs between men and women

**Key point**: Effect modification can be identified when X is
randomized, but Z is merely observed.

## Interaction (Hernán & Robins Definition)

**Interaction** exists when the effect of jointly intervening on both X
and Z differs from what would be predicted by their separate effects.

-   Requires that *both* X and Z are manipulable
-   Formally: Involves potential outcomes under joint interventions
    $Y^{x,z}$
-   Tests for synergy or antagonism between two interventions
-   Example: Combined effect of a drug and exercise program

**Key point**: Interaction requires the ability to intervene on both
variables.

## The Critical Distinction

-   **If X is randomized and Z is randomized**: Effect modification =
    Interaction (identifiable)
-   **If X is randomized but Z is observed**: Effect modification ≠
    Interaction (interaction requires additional assumptions)
-   **If Z is non-manipulable** (e.g., sex, race): Only effect
    modification is conceptually coherent

# Setup

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggdag)
library(dagitty)
library(patchwork)

set.seed(42)

# Set theme for plots
theme_set(theme_minimal(base_size = 12))
```

# Scenario 1: Effect Modification Without True Interaction

## The Causal Structure

In this scenario, we have: - U: An unmeasured confounder - X: A
randomized treatment - Z: An observed variable (confounded by U) - U
modifies the effect of X on Y - Z appears to modify X's effect, but this
is due to confounding

```{r dag1}
# Create DAG for effect modification scenario
dag1 <- dagify(
  Y ~ X + Z + U,  # U modifies X's effect
  Z ~ U,
  exposure = "X",
  outcome = "Y",
  latent = "U",
  coords = list(
    x = c(X = 1, Z = 2, U = 2, Y = 3),
    y = c(X = 2, Z = 3, U = 1, Y = 2)
  )
)

ggdag(dag1, text_size = 4) +
  theme_dag() +
  labs(title = "Effect Modification Due to Confounding",
       subtitle = "U confounds Z→Y and modifies X's effect")
```

## Simulation

```{r sim1}
n <- 10000

# Generate data
simulate_effect_modification <- function(n) {
  # Unmeasured confounder U (continuous)
  U <- rnorm(n, mean = 0, sd = 1)
  
  # Z is caused by U (and correlated with U)
  Z <- rbinom(n, 1, plogis(U))
  
  # X is randomized (independent of U and Z)
  X <- rbinom(n, 1, 0.5)
  
  # Y is affected by X, Z, U, and U modifies X's effect
  # The key: U modifies the effect, not Z!
  Y <- 10 + 
       2 * X +           # Base effect of X
       1 * Z +           # Effect of Z
       1.5 * U +         # Effect of U
       3 * X * U +       # U modifies X's effect (TRUE interaction with U)
       rnorm(n, 0, 1)    # Noise
  
  data.frame(X = X, Z = Z, U = U, Y = Y)
}

data1 <- simulate_effect_modification(n)

# Estimate effect modification by Z (what we observe)
effect_mod_z <- data1 %>%
  group_by(Z) %>%
  summarise(
    effect_of_X = mean(Y[X == 1]) - mean(Y[X == 0]),
    .groups = 'drop'
  )

print("Observed Effect Modification by Z:")
print(effect_mod_z)

# True effect modification by U (what's actually happening)
data1 <- data1 %>%
  mutate(U_tertile = cut(U, breaks = quantile(U, probs = seq(0, 1, 1/3)),
                         include.lowest = TRUE, labels = c("Low", "Mid", "High")))

effect_mod_u <- data1 %>%
  group_by(U_tertile) %>%
  summarise(
    effect_of_X = mean(Y[X == 1]) - mean(Y[X == 0]),
    .groups = 'drop'
  )

print("\nTrue Effect Modification by U:")
print(effect_mod_u)
```

## Visualization

```{r plot1, fig.width=10, fig.height=4}
# Plot 1: Observed effect modification by Z
p1 <- ggplot(data1, aes(x = factor(X), y = Y, fill = factor(Z))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),
                    labels = c("Z = 0", "Z = 1")) +
  labs(title = "Observed: Effect Modification by Z",
       subtitle = "Treatment effect appears different across Z",
       x = "Treatment (X)", y = "Outcome (Y)", fill = "Z") +
  theme(legend.position = "bottom")

# Plot 2: True effect modification by U
p2 <- ggplot(data1, aes(x = factor(X), y = Y, fill = U_tertile)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#999999", "#E69F00", "#56B4E9")) +
  labs(title = "True: Effect Modification by U",
       subtitle = "U is the actual effect modifier",
       x = "Treatment (X)", y = "Outcome (Y)", fill = "U Tertile") +
  theme(legend.position = "bottom")

p1 + p2
```

## Key Insight

We observe effect modification by Z, but Z itself doesn't causally
modify the effect - U does!

To claim **interaction** between X and Z, we'd need to: 1. Intervene on
both X and Z 2. Control for confounding of Z→Y (adjust for U or its
proxies)

Since we cannot intervene on Z (or control for U), we cannot identify
the true X×Z interaction.

# Scenario 2: True Interaction (Both Variables Randomized)

## The Causal Structure

Now both X and Z are randomized interventions, and they truly interact
in producing Y.

```{r dag2}
# Create DAG for true interaction
dag2 <- dagify(
  Y ~ X + Z,  # True interaction
  exposure = c("X", "Z"),
  outcome = "Y",
  coords = list(
    x = c(X = 1, Z = 2, Y = 2.5),
    y = c(X = 1, Z = 1, Y = 2)
  )
)

ggdag(dag2, text_size = 4) +
  theme_dag() +
  labs(title = "True Interaction Between Two Interventions",
       subtitle = "Both X and Z are randomized; they interact mechanistically")
```

## Simulation

```{r sim2}
simulate_true_interaction <- function(n) {
  # Both X and Z are randomized
  X <- rbinom(n, 1, 0.5)
  Z <- rbinom(n, 1, 0.5)
  
  # Y has a true interaction between X and Z
  Y <- 10 + 
       2 * X +           # Main effect of X
       3 * Z +           # Main effect of Z
       5 * X * Z +       # TRUE interaction (synergy)
       rnorm(n, 0, 1)    # Noise
  
  data.frame(X = X, Z = Z, Y = Y)
}

data2 <- simulate_true_interaction(n)

# Estimate effects
effects2 <- data2 %>%
  group_by(Z) %>%
  summarise(
    effect_of_X = mean(Y[X == 1]) - mean(Y[X == 0]),
    .groups = 'drop'
  )

print("Effect of X at each level of Z:")
print(effects2)

# Calculate interaction contrast
interaction_contrast <- effects2$effect_of_X[2] - effects2$effect_of_X[1]
print(paste("\nInteraction contrast:", round(interaction_contrast, 2)))
print("(Difference in effect of X between Z=1 and Z=0)")

# This equals effect modification!
print("\nSince both are randomized:")
print("Effect Modification = Interaction")
```

## Visualization

```{r plot2, fig.width=10, fig.height=5}
# Create means for line plot
means2 <- data2 %>%
  group_by(X, Z) %>%
  summarise(mean_Y = mean(Y), .groups = 'drop')

# Plot 1: Box plots
p3 <- ggplot(data2, aes(x = factor(X), y = Y, fill = factor(Z))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),
                    labels = c("Z = 0", "Z = 1")) +
  labs(title = "True Interaction Between X and Z",
       subtitle = "Both variables randomized",
       x = "Treatment X", y = "Outcome (Y)", fill = "Intervention Z") +
  theme(legend.position = "bottom")

# Plot 2: Interaction plot
p4 <- ggplot(means2, aes(x = factor(X), y = mean_Y, 
                         color = factor(Z), group = factor(Z))) +
  geom_point(size = 4) +
  geom_line(aes(group = factor(Z)), linewidth = 1) +
  scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c("Z = 0", "Z = 1")) +
  labs(title = "Interaction Plot",
       subtitle = "Non-parallel lines indicate interaction",
       x = "Treatment X", y = "Mean Outcome (Y)", color = "Intervention Z") +
  theme(legend.position = "bottom")

p3 + p4
```

## Key Insight

When both X and Z are randomized: - Effect modification IS interaction -
We can make causal claims about synergy - No confounding to worry about

The non-parallel lines show that the effect of X depends on Z (and vice
versa) - this is **true causal synergy**.

# Scenario 3: Comparing Both Scenarios

## Side-by-Side Comparison

```{r comparison, fig.width=12, fig.height=5}
# Add scenario labels
data1_labeled <- data1 %>%
  mutate(Scenario = "Scenario 1: X randomized, Z observed")

data2_labeled <- data2 %>%
  mutate(Scenario = "Scenario 2: Both randomized") %>%
  select(X, Z, Y, Scenario)

# Combine
combined <- bind_rows(
  data1_labeled %>% select(X, Z, Y, Scenario),
  data2_labeled
)

# Calculate effects for each scenario
effects_comparison <- combined %>%
  group_by(Scenario, Z) %>%
  summarise(
    effect_of_X = mean(Y[X == 1]) - mean(Y[X == 0]),
    .groups = 'drop'
  )

# Plot
ggplot(effects_comparison, aes(x = factor(Z), y = effect_of_X, 
                                fill = Scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  labs(title = "Effect of X by Z: Both Scenarios",
       subtitle = "Effect modification appears in both, but only Scenario 2 has identifiable interaction",
       x = "Z", y = "Effect of X on Y", fill = "") +
  theme(legend.position = "bottom")
```

## Statistical Models

Let's fit models to both scenarios and interpret the interaction terms.

```{r models}
# Scenario 1: Observed effect modification
model1 <- lm(Y ~ X * Z, data = data1)
cat("SCENARIO 1: X randomized, Z observed\n")
cat("=====================================\n")
summary(model1)$coefficients

cat("\nInterpretation:\n")
cat("- X:Z coefficient captures observed effect modification\n")
cat("- Does NOT represent true causal interaction (confounded by U)\n")
cat("- To identify interaction, would need to control for U\n\n")

# Scenario 2: True interaction
model2 <- lm(Y ~ X * Z, data = data2)
cat("SCENARIO 2: Both randomized\n")
cat("===========================\n")
summary(model2)$coefficients

cat("\nInterpretation:\n")
cat("- X:Z coefficient represents TRUE causal interaction\n")
cat("- Effect modification = Interaction (both identifiable)\n")
cat("- No confounding concerns\n")
```

# Summary Table

```{r summary_table}
summary_df <- data.frame(
  Concept = c("Effect Modification", "Interaction (H&R)"),
  Definition = c(
    "Effect of X varies across levels of Z",
    "Effect of joint intervention on X and Z shows synergy/antagonism"
  ),
  Requirements = c(
    "X randomized or conditional exchangeability for X",
    "Both X and Z randomized or conditional exchangeability for both"
  ),
  Z_Status = c(
    "Can be observed (non-manipulable)",
    "Must be manipulable"
  ),
  Example = c(
    "Drug effect differs by sex",
    "Drug + exercise show synergy"
  ),
  Identifiable_When = c(
    "X is randomized",
    "Both X and Z are randomized (or strong assumptions)"
  )
)

knitr::kable(summary_df, caption = "Comparison of Effect Modification and Interaction")
```

# Key Takeaways

1.  **Effect modification and interaction are not properties of the
    data-generating process alone** - they depend on what causal
    question we're asking and what we can identify.

2.  **Effect modification** (observed) can exist without true
    **interaction** (causal):

    -   When Z is confounded, we may observe effect modification by Z
    -   But the true mechanism may involve an unmeasured confounder U
    -   This is why effect modification ≠ interaction when Z isn't
        randomized

3.  **When both variables are randomized**, effect modification =
    interaction:

    -   No confounding to worry about
    -   Statistical interaction term has causal interpretation
    -   We can make claims about synergy/antagonism

4.  **The statistical tool (interaction term in regression) is the
    same** in both cases, but the **causal interpretation differs
    fundamentally**:

    -   With one randomized: Describes how observed effect varies
        (possibly confounded)
    -   With both randomized: Describes true causal synergy

5.  **Interaction requires manipulability** of both variables:

    -   Can't have H&R "interaction" with sex, race, genetics
    -   Can only have effect modification in these cases
    -   This is a conceptual distinction, not just a statistical one

# References

-   Hernán MA, Robins JM. *Causal Inference: What If.* Boca Raton:
    Chapman & Hall/CRC, 2020.
-   VanderWeele TJ, Robins JM. The identification of synergism in the
    sufficient-component-cause framework. *Epidemiology.*
    2007;18(3):329-339.
-   Rothman KJ, Greenland S, Lash TL. *Modern Epidemiology.* 3rd ed.
    Philadelphia: Lippincott Williams & Wilkins, 2008.

------------------------------------------------------------------------

**Session Info**

```{r session}
sessionInfo()
```
