---
title: "Understanding Variance and Covariance"
format: html
editor: visual
---

## Introduction

Variance and covariance are fundamental concepts in statistics that help us understand how data varies and how variables relate to each other. In this document, we'll explore both concepts using R code and visualizations.

## Variance

**Variance** measures how spread out a set of numbers is from their mean (average). It tells us how much the data points differ from the average value.

### Mathematical Definition

The variance of a dataset is calculated as:

$$\text{Var}(X) = \frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})^2$$

Where: - $x_i$ are individual data points - $\bar{x}$ is the mean of the data - $n$ is the number of observations

### Example: Calculating Variance Step-by-Step

Let's create a simple dataset and calculate its variance, breaking down each step:

```{r}
# Create a sample dataset
data <- c(2, 4, 4, 4, 5, 5, 7, 9)

# Calculate the mean
mean_value <- mean(data)

# Create a detailed table showing each step
variance_table <- data.frame(
  i = 1:length(data),
  x_i = data,
  mean_x = mean_value,
  deviation = data - mean_value,
  squared_deviation = (data - mean_value)^2
)

colnames(variance_table) <- c("i", "x_i", "mean(X)", "x_i - mean", "(x_i - mean)²")

library(knitr)
kable(variance_table, digits = 2, caption = "Step-by-Step Variance Calculation")

# Calculate variance
variance_value <- sum(variance_table$`(x_i - mean)²`) / length(data)
cat("\nSum of squared deviations:", sum(variance_table$`(x_i - mean)²`), "\n")
cat("Variance = Sum / n = ", sum(variance_table$`(x_i - mean)²`), "/", length(data), "=", variance_value, "\n")
```

**Understanding the table:** - Column 1: Index of each observation - Column 2: The actual data values - Column 3: The mean (constant for all rows) - Column 4: How far each point is from the mean (can be positive or negative) - Column 5: The squared distance from the mean (always positive)

Note: R's `var()` function uses $n-1$ in the denominator (sample variance), while we're using $n$ (population variance) here for simplicity.

### Visualizing the Variance Calculation

Now let's see this graphically. First, we'll plot the data points and the mean:

```{r}
#| fig-width: 10
#| fig-height: 6

library(ggplot2)

# Create data frame for plotting
df_variance <- data.frame(
  i = 1:length(data),
  x_i = data
)

# Plot 1: Data points and mean with deviation lines
p1 <- ggplot(df_variance, aes(x = i, y = x_i)) +
  geom_hline(yintercept = mean_value, color = "red", size = 1.2, linetype = "dashed") +
  geom_segment(aes(xend = i, yend = mean_value), 
               color = "blue", alpha = 0.6, size = 1) +
  geom_point(size = 4, color = "darkblue") +
  annotate("text", x = 8.5, y = mean_value + 0.3, 
           label = paste("Mean =", round(mean_value, 2)), 
           color = "red", size = 5, hjust = 1) +
  labs(
    title = "Deviations from the Mean",
    subtitle = "Blue lines show how far each point is from the mean (red line)",
    x = "Observation Index (i)",
    y = "Value (x_i)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p1)
```

**Key Insight:** The blue lines represent the deviations $(x_i - \bar{x})$. Some are positive (above the mean) and some are negative (below the mean).

### Why Do We Square the Deviations?

If we just averaged the deviations without squaring, positive and negative deviations would cancel out, always giving us zero! Let's visualize this:

```{r}
#| fig-width: 12
#| fig-height: 5

library(gridExtra)

# Show why squaring matters
deviations <- data - mean_value

# Create visualization of squares
df_squares <- data.frame(
  i = 1:length(data),
  x_i = data,
  deviation = deviations,
  abs_deviation = abs(deviations)
)

# Plot showing signed deviations
p2 <- ggplot(df_squares, aes(x = i, y = deviation)) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  geom_col(aes(fill = deviation > 0), alpha = 0.7) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(
    title = "Deviations (with sign)",
    subtitle = paste("Sum =", round(sum(deviations), 2)),
    x = "Observation (i)",
    y = "Deviation from Mean"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot showing squared deviations
p3 <- ggplot(df_squares, aes(x = i, y = deviation^2)) +
  geom_col(fill = "purple", alpha = 0.7) +
  labs(
    title = "Squared Deviations",
    subtitle = paste("Sum =", round(sum(deviations^2), 2)),
    x = "Observation (i)",
    y = "Squared Deviation"
  ) +
  theme_minimal()

grid.arrange(p2, p3, ncol = 2)
```

**Why squaring works:** - Left plot: Positive and negative deviations cancel out (sum ≈ 0) - Right plot: Squaring makes all deviations positive, so they add up meaningfully - Larger deviations contribute more to variance (because squaring amplifies them)

### Geometric Interpretation: Squares!

We can actually visualize the "squared" deviations as literal squares:

```{r}
#| fig-width: 10
#| fig-height: 8

# Create a plot showing actual geometric squares
df_plot <- df_variance
df_plot$deviation <- abs(data - mean_value)

# Create square coordinates for each point
create_square <- function(x_center, y_center, side_length) {
  data.frame(
    x = x_center + c(-side_length/2, side_length/2, side_length/2, -side_length/2, -side_length/2),
    y = y_center + c(-side_length/2, -side_length/2, side_length/2, side_length/2, -side_length/2)
  )
}

# Create squares for visualization (scaled for visibility)
squares_list <- lapply(1:nrow(df_plot), function(i) {
  sq <- create_square(
    x_center = df_plot$i[i], 
    y_center = mean_value - df_plot$deviation[i]/2,
    side_length = df_plot$deviation[i]
  )
  sq$group <- i
  sq
})

squares_df <- do.call(rbind, squares_list)

ggplot() +
  geom_polygon(data = squares_df, 
               aes(x = x, y = y, group = group), 
               fill = "lightblue", color = "darkblue", alpha = 0.5, size = 0.8) +
  geom_hline(yintercept = mean_value, color = "red", size = 1.2, linetype = "dashed") +
  geom_point(data = df_plot, aes(x = i, y = x_i), size = 4, color = "darkblue") +
  annotate("text", x = 8.5, y = mean_value + 0.3, 
           label = paste("Mean =", round(mean_value, 2)), 
           color = "red", size = 5, hjust = 1) +
  labs(
    title = "Geometric Interpretation: Squared Deviations as Actual Squares",
    subtitle = "Each square has an area equal to (deviation)². Variance is the average of these areas.",
    x = "Observation Index (i)",
    y = "Value"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, face = "bold"))
```

**Geometric Insight:** Each blue square has an area equal to $(x_i - \bar{x})^2$. The variance is literally the average area of these squares! Points farther from the mean have larger squares, contributing more to the overall variance.

### Visualizing Variance

Let's visualize how variance represents spread:

```{r}
#| fig-width: 10
#| fig-height: 6

library(ggplot2)

# Create two datasets with different variances
set.seed(42)
low_variance <- rnorm(100, mean = 50, sd = 5)
high_variance <- rnorm(100, mean = 50, sd = 15)

# Combine into a data frame
df <- data.frame(
  value = c(low_variance, high_variance),
  group = rep(c("Low Variance", "High Variance"), each = 100)
)

# Calculate variances
low_var <- var(low_variance)
high_var <- var(high_variance)

# Create visualization
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "black", size = 1) +
  labs(
    title = "Comparing Low and High Variance Distributions",
    subtitle = sprintf("Low Variance: %.2f | High Variance: %.2f", low_var, high_var),
    x = "Value",
    y = "Frequency",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Key Insight:** The high variance group is much more spread out from the mean (dashed line), while the low variance group clusters tightly around it.

## Covariance

**Covariance** measures how two variables change together. It tells us whether increases in one variable tend to correspond with increases (positive covariance) or decreases (negative covariance) in another variable.

### Mathematical Definition

The covariance between two variables X and Y is:

$$\text{Cov}(X,Y) = \frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})$$

### Example: Calculating Covariance

```{r}
# Create two related variables
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
y <- c(2, 4, 5, 7, 8, 10, 11, 13, 14, 16)

# Calculate means
mean_x <- mean(x)
mean_y <- mean(y)

# Calculate covariance manually
cov_manual <- sum((x - mean_x) * (y - mean_y)) / length(x)
cat("Covariance (manual):", cov_manual, "\n")

# Calculate covariance using R
cov_r <- cov(x, y)
cat("Covariance (R function):", cov_r, "\n")
```

### Visualizing Covariance

Let's visualize different types of covariance relationships:

```{r}
#| fig-width: 12
#| fig-height: 4

set.seed(123)
n <- 100

# Positive covariance
x_pos <- rnorm(n)
y_pos <- x_pos + rnorm(n, sd = 0.5)

# Negative covariance
x_neg <- rnorm(n)
y_neg <- -x_neg + rnorm(n, sd = 0.5)

# No covariance (independent)
x_none <- rnorm(n)
y_none <- rnorm(n)

# Create data frame
df_cov <- data.frame(
  x = c(x_pos, x_neg, x_none),
  y = c(y_pos, y_neg, y_none),
  type = rep(c(
    sprintf("Positive Covariance\n(%.2f)", cov(x_pos, y_pos)),
    sprintf("Negative Covariance\n(%.2f)", cov(x_neg, y_neg)),
    sprintf("No Covariance\n(%.2f)", cov(x_none, y_none))
  ), each = n)
)

# Plot
ggplot(df_cov, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  facet_wrap(~ type, nrow = 1) +
  labs(
    title = "Three Types of Covariance Relationships",
    x = "Variable X",
    y = "Variable Y"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 10, face = "bold"))
```

**Interpretation:** - **Positive Covariance:** When X increases, Y tends to increase - **Negative Covariance:** When X increases, Y tends to decrease\
- **Zero Covariance:** X and Y are independent (no linear relationship)

## Practical Example: Heights and Weights

Let's look at a real-world example using simulated height and weight data:

```{r}
#| fig-width: 10
#| fig-height: 6

# Simulate height and weight data
set.seed(456)
n_people <- 200

heights <- rnorm(n_people, mean = 170, sd = 10)  # cm
weights <- 0.5 * heights + rnorm(n_people, mean = -15, sd = 5)  # kg

# Calculate statistics
height_var <- var(heights)
weight_var <- var(weights)
height_weight_cov <- cov(heights, weights)
correlation <- cor(heights, weights)

# Create scatter plot
df_people <- data.frame(heights, weights)

ggplot(df_people, aes(x = heights, y = weights)) +
  geom_point(alpha = 0.6, color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Relationship Between Height and Weight",
    subtitle = sprintf(
      "Variance(Height): %.2f | Variance(Weight): %.2f | Covariance: %.2f | Correlation: %.2f",
      height_var, weight_var, height_weight_cov, correlation
    ),
    x = "Height (cm)",
    y = "Weight (kg)"
  ) +
  theme_minimal()
```

**Key Takeaway:** Height and weight have positive covariance, meaning taller people tend to weigh more.

## Variance vs. Covariance: Key Differences

| Aspect | Variance | Covariance |
|----|----|----|
| **Variables** | One variable | Two variables |
| **Measures** | Spread of data | Relationship between variables |
| **Range** | Always positive (≥ 0) | Can be positive, negative, or zero |
| **Units** | Squared units of variable | Product of the two variables' units |

## Covariance Matrix

When working with multiple variables, we often use a **covariance matrix** that shows all pairwise covariances:

```{r}
# Create a dataset with multiple variables
set.seed(789)
df_multi <- data.frame(
  var1 = rnorm(50, mean = 10, sd = 2),
  var2 = rnorm(50, mean = 20, sd = 3),
  var3 = rnorm(50, mean = 30, sd = 4)
)

# Make var2 correlated with var1
df_multi$var2 <- df_multi$var1 * 1.5 + rnorm(50, sd = 2)

# Calculate covariance matrix
cov_matrix <- cov(df_multi)
print("Covariance Matrix:")
print(round(cov_matrix, 2))

# Visualize the covariance matrix
library(reshape2)
cov_melted <- melt(cov_matrix)

ggplot(cov_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "white", size = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Covariance Matrix Heatmap",
    x = "",
    y = "",
    fill = "Covariance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The diagonal elements show the variance of each variable, while off-diagonal elements show covariances between pairs of variables.

## Summary

-   **Variance** tells us how spread out a single variable's values are
    -   Higher variance = more spread out
    -   Lower variance = more clustered around the mean
-   **Covariance** tells us how two variables move together
    -   Positive = variables increase together
    -   Negative = one increases while the other decreases
    -   Zero = no linear relationship
-   Both are fundamental for understanding data distributions and relationships in statistics and machine learning!

## Further Reading

-   Correlation is a standardized version of covariance (ranging from -1 to 1)
-   Covariance matrices are essential in Principal Component Analysis (PCA)
-   Understanding these concepts is crucial for regression analysis and portfolio theory
